smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-08-07 manuscript/manuscript.qmd
added lay summary, changed to jamia citation style, minor modifications to manuscript (mostly shortening and wording)

---
title: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases"

@@

editor: visual
bibliography: references.bib
csl: computer-methods-and-programs-in-biomedicine

@@

[**Word count:**]{.underline} 2,095 words/2,000 words

[**Tables:**]{.underline} 2**/2**

[**Figures:**]{.underline} 3**/3**

[**Supplementary material:**]{.underline}

@@

[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All R code, materials and depedency information can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>. The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from <https://janickweberpals.gitlab-pages.partners.org/smdi/>.

@@

148 words/150 words

@@

Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR). While analytic approaches exist (e.g., multiple imputation), assumption on underlying missingness patterns and mechanisms have to hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

We developed the smdi (structural missing data diagnostics) R package based on structural assumptions of common missing data mechanisms in real-world data and results of a previous simulation study using a US EHR-claims database linkage.

@@

smdi enables users to run principled missing data investigations on partially observed confounders including descriptive analyses and diagnostics on potentially underlying missingness patterns and mechanisms.

@@

smdi can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies. smdi is freely available seamlessly integrates to the ecosystem of R packages for healthcare database analytics.

@@

-   ...
:::

@@

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

In light of these limitations, we have recently developed and evaluated a principled approach with multiple missing data diagnostics in an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated daily R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The workflow is generally categorized into descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these can be interpreted. Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>). For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs along with a corresponding range of minimum and maximum AMSDs observed. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. For a graphical visualization of this, the function returns a plot illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and p-value for each each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this POC and run the diagnostics for each POC independently.

@@

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the the regression type, a user needs to specify the column name that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not. The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

@@

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in real-world databases and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data. By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, \[@weberpals2023\]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.

The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important that diagnostic results are complemented with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.

@@

The smdi package is a powerful and convenient tool to implement and carry out principled missing data diagnostics in RWE studies and improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset. Through its design it seamlessly integrates with other R packages commonly used for clinical reporting [@gtsummary; @gt; @mice; @naniar; @ggplot2].

>>>>

---
title: "smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies"

@@

editor: visual
bibliography: references.bib
csl: jamia.csl

@@

[**Word count:**]{.underline} 2,070 words/2,000 words

[**Tables:**]{.underline} 2/2

[**Figures:**]{.underline} 3/3

[**Supplementary material:**]{.underline} None

@@

[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>. The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.

@@

147 words/150 words

@@

Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR). While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms have to hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

We developed the smdi (structural missing data diagnostics) R package based on results of a previous simulation study which considered structural assumptions of common missing data mechanisms in real-world data.

@@

smdi enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.

@@

The smdi R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records). Hence, confounders and prognostic factors, which need to be balanced between treatment groups in real-world evidence studies, are usually not available for all patients at all necessary time points which leads to missing data. If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used. Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold. Here, we present the smdi R package, which enables researchers to perform such principled missing data investigations on partially observed confounders. The smdi package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on available data.
:::

@@

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

In light of these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted. Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. For a graphical visualization of this, the function returns a plot for each POC illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.

@@

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the the regression type (`model` parameter), a user needs to specify the column name containing the outcome using the `form_lhs` parameter (e.g., `Surv(eventtime, status)` in case of a Cox model). The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

@@

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in real-world databases and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data. By cross-checking the resulting diagnostic parameters to expected estimates ( @fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.

The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.

@@

The smdi package is a powerful and convenient tool to implement and carry out principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.
